---
title: "Winnipeg Open Datathon — Import & Diagnostics"
author: "Myo Minn Oo & Yong Jin"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    highlight: tango
    df_print: default
    code_folding: hide
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE
)
```

---

# Section 1: Data Import, Cleaning, And Processing

This document covers the full pipeline for importing, caching, and diagnosing
11 datasets from the [Winnipeg Open Data Portal](https://data.winnipeg.ca).
The datasets span property assessment, public health, transit, land use, and
census data. The ultimate goal is to merge them into a parcel-level master
dataset anchored on **Assessment Parcels**.

---

# 1. API Configuration

Each dataset is referenced by its Socrata resource ID. Per-dataset row limits
are defined separately so they can be adjusted without touching the fetch logic.
`NULL` means no limit — `RSocrata::read.socrata()` will auto-paginate and
return all rows.

Note: Two datasets — **Transit Pass-ups** and **Short Term Rentals** — have no
neighbourhood identifier and will require a manual spatial join later.

```{r api-config}
api_list <- list(
  assessment_parcels        = "https://data.winnipeg.ca/resource/d4mq-wa44.json",
  parks_open_space          = "https://data.winnipeg.ca/resource/tx3d-pfxq.json",
  substance_use             = "https://data.winnipeg.ca/resource/6x82-bz5y.json",
  naloxone_administrations  = "https://data.winnipeg.ca/resource/qd6b-q49i.json",
  transit_passups           = "https://data.winnipeg.ca/resource/mer2-irmb.json",  # no neighbourhood
  census_households         = "https://data.winnipeg.ca/resource/nmk5-uwfw.json",
  requests_311              = "https://data.winnipeg.ca/resource/u7f6-5326.json",
  higher_poverty_areas      = "https://data.winnipeg.ca/resource/gg4w-peq6.json",
  trade_permits             = "https://data.winnipeg.ca/resource/urbd-qygv.json",
  rooming_house_enforcement = "https://data.winnipeg.ca/resource/vk2f-xwp7.json",
  short_term_rentals        = "https://data.winnipeg.ca/resource/74hr-f8ai.json"   # no neighbourhood
)

api_limits <- list(
  assessment_parcels        = 50000,
  parks_open_space          = NULL,
  substance_use             = NULL,
  naloxone_administrations  = NULL,
  transit_passups           = 50000,
  census_households         = NULL,
  requests_311              = 50000,
  higher_poverty_areas      = NULL,
  trade_permits             = 50000,
  rooming_house_enforcement = NULL,
  short_term_rentals        = 5000
)
```

---

# 2. Fetch Functions

With Claude's help, we have created three modular functions handle the import pipeline:

- `build_url()` — appends `?$limit=N` to a URL when a limit is specified
- `fetch_socrata()` — fetches a single dataset with `tryCatch` error handling;
  returns `NULL` on failure so the rest of the pipeline continues uninterrupted
- `fetch_all_socrata()` — orchestrates the full fetch using `purrr::map2()` to
  pair each URL with its limit, then reports any failures
  
  
```{r fetch-functions}
# Module 1: Build URL with optional row limit ----
build_url <- function(url, limit = NULL) {
  if (!base::is.null(limit)) {
    base::paste0(url, "?$limit=", limit)
  } else {
    url
  }
}

# Module 2: Fetch a single dataset with error handling ----
fetch_socrata <- function(url, limit = NULL) {
  final_url <- build_url(url, limit)
  base::message("Fetching: ", final_url)
  tryCatch(
    expr    = RSocrata::read.socrata(final_url),
    error   = \(e) {
      base::message("ERROR fetching ", final_url, "\n  → ", base::conditionMessage(e))
      NULL
    },
    warning = \(w) {
      base::message("WARNING fetching ", final_url, "\n  → ", base::conditionMessage(w))
      RSocrata::read.socrata(final_url)
    }
  )
}

# Module 3: Report failed datasets ----
report_failures <- function(data_list) {
  failed <- base::names(purrr::keep(data_list, base::is.null))
  if (base::length(failed) > 0) {
    base::message("\nFailed to load: ", base::paste(failed, collapse = ", "))
  } else {
    base::message("\nAll datasets loaded successfully.")
  }
  base::invisible(failed)
}

# Module 4: Fetch all datasets with per-dataset limits ----
fetch_all_socrata <- function(api_list, api_limits = NULL) {
  if (base::is.null(api_limits)) {
    data_list <- purrr::map(api_list, fetch_socrata)
  } else {
    data_list <- purrr::map2(
      api_list,
      api_limits,
      \(url, lim) fetch_socrata(url, limit = lim)
    )
  }
  report_failures(data_list)
  data_list
}
```

---

# 3. Import Data

Data is loaded from a local cache (`data/data_raw_list.RData`) if it exists,
avoiding unnecessary repeat API calls. If no cache is found, data is fetched
fresh from the API and saved for future sessions.
```{r import-data}
cache_path <- "data/data_raw_list.RData"

if (base::file.exists(cache_path)) {
  base::message("Loading from cache: ", cache_path)
  base::load(cache_path)
} else {
  base::message("Cache not found — fetching from API...")
  data_raw_list <- fetch_all_socrata(api_list, api_limits)
  base::save(data_raw_list, file = cache_path)
}
```

---

# 4. Structure Overview

Quick check of rows and columns per dataset to verify the pull was complete.
```{r structure-overview}
dim_summary <- purrr::imap_dfr(data_raw_list, \(df, name) {
  tibble::tibble(
    dataset = name,
    rows    = base::nrow(df),
    cols    = base::ncol(df),
    loaded  = !base::is.null(df)
  )
})

dim_summary |> 
  make_dt()
```

---

# 5. Join Key Scan

Scan all datasets for columns likely to serve as merge keys — neighbourhood,
ward, and ID fields.
```{r join-key-scan}
join_keys <- purrr::imap_dfr(data_raw_list, \(df, name) {
  if (base::is.null(df)) return(NULL)
  matched <- base::grep(
    pattern     = "neighbourhood|neigh|_id|ward",
    x           = base::names(df),
    value       = TRUE,
    ignore.case = TRUE
  )
  if (base::length(matched) == 0) return(NULL)
  tibble::tibble(
    dataset   = name,
    join_keys = base::paste(matched, collapse = ", ")
  )
})

join_keys |> 
  make_dt()
```

---

# 6. Column Diagnostics

The `check_all()` function runs a comprehensive column-level audit across every
dataset. For each column it reports:

- **n_total / n_valid / n_missing** — observation counts
- **pct_missing** — percentage of missing or blank values
- **min / max / mean / median / sd** — for numeric columns
- **min / max** (date range) — for date columns
- **n_unique / categories** — top categories with counts for character columns

---

## 6.1 Missing Values Summary

Columns with any missing values, ranked by severity.
```{r missing-values}
# missing_summary <- diagnostics |>
#   dplyr::filter(pct_missing > 0) |>
#   dplyr::select(dataset, column, type, n_total, n_missing, pct_missing) |>
#   dplyr::arrange(dplyr::desc(pct_missing)) 
# 
# missing_summary |> 
#   make_dt()
```
```{r missing-flag}
# Columns with > 50% missing — candidates for exclusion ----
# missing_summary |>
#   dplyr::filter(pct_missing > 50) |>
#   dplyr::select(dataset, column, pct_missing) |> 
#   make_dt()
```

---

## 6.2 Numeric Column Statistics

Summary statistics for all numeric and integer columns across every dataset.
```{r numeric-stats}
# diagnostics |>
#   dplyr::filter(type %in% c("numeric", "integer")) |>
#   dplyr::select(dataset, column, min, max, mean, median, sd, n_missing, pct_missing) |> 
#   make_dt()
```

---

## 6.3 Categorical Column Valid Values

Valid categories and their frequencies for all character columns. Columns with
more than 10 unique values show the top 10 plus a count of remaining categories.
```{r categorical-values}
# diagnostics |>
#   dplyr::filter(type == "character", !base::is.na(categories)) |>
#   dplyr::select(dataset, column, n_unique, categories) |> 
#   make_dt()
```

---

## 6.4 Date Column Ranges

Date range (min → max) for all date and datetime columns — useful for
understanding temporal coverage before merging.
```{r date-ranges}
# diagnostics |>
#   dplyr::filter(type %in% c("POSIXct", "POSIXlt", "Date")) |>
#   dplyr::select(dataset, column, min, max, n_unique) |> 
#   make_dt()
```


---

# 8. Session Info
```{r session-info}
utils::sessionInfo()
```



# Section 2: Domain Structure Factor Analysis

The Winnipeg Neighbourhood Liveability Index (WNLI) is a composite score that
summarises the overall quality of life across Winnipeg neighbourhoods using
publicly available open data from the
[Winnipeg Open Data Portal](https://data.winnipeg.ca). The index aggregates
11 datasets spanning property assessment, public health, green space, municipal
services, housing vulnerability, and short-term rentals into a single 0–100
score per neighbourhood.

The index is designed to be **transparent**, **reproducible**, and
**comparable across neighbourhoods** — three properties that guided every
methodological decision documented here.

---

# 1. Common Approaches to Liveability Indices

There are three established methods for constructing composite liveability
indices. Each involves tradeoffs between simplicity, statistical rigour, and
communicability.

---

## 1.1 Equal-Weight Composite Scoring

The simplest approach. Each indicator is normalized to a common scale
(typically 0–100 or a z-score), then all indicators are averaged with equal
weight to produce a final score.

**Used by:** The Economist Intelligence Unit Global Liveability Index,
MERCER Quality of Living Survey.

**Strengths:**

- Fully transparent — every indicator contributes the same amount
- Easy to explain to non-technical audiences
- Simple to update as new data becomes available

**Weaknesses:**

- Implicitly assumes all dimensions of liveability are equally important
- Sensitive to the number of indicators per domain — domains with more
  indicators get more influence
- Does not account for statistical redundancy between correlated indicators

---

## 1.2 Domain-Weighted Composite Scoring

Indicators are grouped into thematic **domains** (e.g. safety, housing, green
space). Scores are computed within each domain, then domains are combined with
explicit weights — either equal across domains, or set by expert judgment or
community consultation.

**Used by:** The Canadian Index of Wellbeing, the UN Human Development Index,
most municipal and regional liveability indices.

**Strengths:**

- Balances domains regardless of how many indicators each contains
- Weights can be adjusted to reflect local priorities
- Results are interpretable at the domain level, not just the composite

**Weaknesses:**

- Domain weights are subjective unless derived from community input or research
- Requires clear theoretical justification for what belongs in each domain

> **This is the approach adopted for the WNLI.** Equal weights are applied
> across six domains in the first iteration. Sensitivity analysis will test
> alternative weighting schemes.

---

## 1.3 Principal Component Analysis / Factor Analysis

A data-driven approach where the statistical structure of the indicators
determines how they are grouped and weighted. PCA identifies linear combinations
of indicators that explain the most variance, while factor analysis uncovers
latent constructs underlying the observed variables.

**Used by:** Academic liveability research, neighbourhood deprivation indices,
Statistics Canada's Material and Social Deprivation Index.

**Strengths:**

- Weights are derived from the data, not subjective judgment
- Automatically handles correlated indicators by consolidating them
- Defensible in peer-reviewed or technical contexts

**Weaknesses:**

- Results are difficult to communicate to general audiences
- Component loadings can be unstable with small samples or high missingness
- Less practical for iterative updates as new data arrives

---

# 2. Normalization Methods

Before indicators can be combined into a composite score, they must be placed
on a common scale. The three most common normalization methods are:

| Method | Formula | Range | Best for |
|---|---|---|---|
| **Min-max** | (x − min) / (max − min) | 0 to 1 | Interpretable scores bounded between 0 and 100 |
| **Z-score** | (x − mean) / sd | Unbounded | Comparing deviations from the city average |
| **Rank percentile** | rank(x) / n | 0 to 1 | Robust to outliers; less sensitive to extreme values |

The WNLI uses **min-max normalization** scaled to 0–100. This produces scores
that are immediately interpretable — a neighbourhood scoring 75 performs better
than 75% of the possible range on that indicator. Z-scores will be computed
as a robustness check.

---

# 3. Domain Structure

The WNLI organises its indicators into **six domains**, each reflecting a
distinct dimension of neighbourhood liveability. The domain structure is
informed by the Canadian Index of Wellbeing framework and adapted to the
variables available in Winnipeg's open data.

---

## Domain 1 — Housing & Property

Reflects the quality, size, age, and value of the residential housing stock
in each neighbourhood.

| Indicator | Source | Direction |
|---|---|---|
| Median assessed property value | Assessment Parcels | ↑ Higher = better |
| Median living area (sq ft) | Assessment Parcels | ↑ Higher = better |
| Median housing age | Assessment Parcels | ↓ Older = worse |
| Total dwelling units | Assessment Parcels | ↑ Higher = better |
| Assessment value growth rate | Assessment Parcels | ↑ Higher = better |
| % residential land use | Assessment Parcels | ↑ Higher = better |

---

## Domain 2 — Green Space & Environment

Reflects access to parks and outdoor amenity within each neighbourhood.

| Indicator | Source | Direction |
|---|---|---|
| Parks per 1,000 residents | Parks & Open Space | ↑ Higher = better |
| Park area (ha) per 1,000 residents | Parks & Open Space | ↑ Higher = better |
| Active park & open space area (ha) | Parks & Open Space | ↑ Higher = better |
| Water feature area (ha) | Parks & Open Space | ↑ Higher = better |

---

## Domain 3 — Public Safety & Health

Reflects the burden of substance use-related incidents and opioid overdoses
within each neighbourhood during the 2021–2023 index window.

| Indicator | Source | Direction |
|---|---|---|
| Substance use incidents per 1,000 residents | Substance Use | ↓ Higher = worse |
| Opioid incidents per 1,000 residents | Substance Use | ↓ Higher = worse |
| Crystal meth incidents per 1,000 residents | Substance Use | ↓ Higher = worse |
| Naloxone administrations per 1,000 residents | Naloxone Administrations | ↓ Higher = worse |
| Average naloxone doses per incident | Naloxone Administrations | ↓ Higher = worse |

---

## Domain 4 — Municipal Services & Activity

Reflects both the demand for municipal services and the level of economic
activity and investment in each neighbourhood.

| Indicator | Source | Direction |
|---|---|---|
| 311 requests per 1,000 residents | 311 Requests | Ambiguous — see note |
| 311 request resolution rate (closed/total) | 311 Requests | ↑ Higher = better |
| Trade permits per 1,000 residents | Trade Permits | ↑ Higher = better |
| Trade permit issuance rate (issued/total) | Trade Permits | ↑ Higher = better |

> **Note on 311 requests:** A high volume of 311 requests can indicate either
> an engaged, civically active neighbourhood or a distressed neighbourhood with
> high service needs. The raw count is therefore treated as **ambiguous** and
> excluded from scoring. Only the **resolution rate** is included as a positive
> indicator of service delivery quality.

---

## Domain 5 — Housing Vulnerability

Reflects the concentration of poverty and housing precarity within each
neighbourhood, using both the Market Basket Measure and rooming house
enforcement data.

| Indicator | Source | Direction |
|---|---|---|
| Poverty rate (population in MBM / total) | Higher Poverty Areas | ↓ Higher = worse |
| High poverty area designation | Higher Poverty Areas | ↓ Present = worse |
| Rooming house complaints per neighbourhood | Rooming House Enforcement | ↓ Higher = worse |
| Rooming house enforcement completion rate | Rooming House Enforcement | ↑ Higher = better |

---

## Domain 6 — Housing Market Pressure

Reflects the presence of short-term rental activity, particularly
investor-owned non-primary rentals which are associated with reduced housing
affordability.

| Indicator | Source | Direction |
|---|---|---|
| Short-term rentals per ward | Short-Term Rentals | Ambiguous — see note |
| % non-primary (investor-owned) STRs | Short-Term Rentals | ↓ Higher = worse |

> **Note:** Short-term rental data joins at the **electoral ward level** only,
> as no neighbourhood identifier is available in this dataset. Ward-level values
> are applied uniformly to all parcels within a ward, which introduces
> measurement imprecision at the neighbourhood level.

---

# 4. Composite Score Construction

The final WNLI score is computed in three steps:

**Step 1 — Normalize each indicator** to a 0–1 scale using min-max
normalization. Negative indicators (where higher raw values indicate worse
outcomes) are inverted so that **higher always means more liveable** across
all indicators.

**Step 2 — Average within domains.** Each domain score is the unweighted mean
of its constituent normalized indicators. Missing values are excluded from the
average rather than imputed, so neighbourhoods with fewer data sources still
receive partial scores.

**Step 3 — Average across domains.** The final composite score is the
unweighted mean of all six domain scores, multiplied by 100 to produce a
0–100 scale.

$$
\text{WNLI}_i = \frac{1}{6} \sum_{d=1}^{6} \text{Domain}_{d,i} \times 100
$$

Where $i$ indexes neighbourhoods and $d$ indexes the six domains.

---

# 5. Key Methodological Decisions

| Decision | Choice made | Rationale |
|---|---|---|
| **Index level** | Neighbourhood | Parcel-level data aggregated to neighbourhood before scoring |
| **Normalization** | Min-max (0–100) | Interpretable, bounded, easy to communicate |
| **Domain weights** | Equal (1/6 each) | Transparent starting point; sensitivity analysis will test alternatives |
| **Missing values** | Excluded from average | Avoids imputation bias; partial scores still produced |
| **Negative indicators** | Inverted before scoring | Higher always = more liveable across all indicators |
| **Time window** | 2021–2023 | Tightest common window across all event-level datasets |

---

# 6. Limitations

**Data availability.** Not all neighbourhoods have records in every dataset.
The Higher Poverty Areas dataset covers only ~60 designated high-poverty
neighbourhoods; Domain 5 scores will be missing for all other neighbourhoods
unless this is treated as "not high poverty" (a valid but explicit assumption).

**Neighbourhood name standardisation.** Several datasets use different naming
conventions for the same neighbourhood. Fuzzy matching and a manual remap table
were used to standardise keys before joining, but residual mismatches may remain.

**Ecological fallacy.** All scores are neighbourhood-level averages. They
describe the neighbourhood as a whole and should not be used to make inferences
about individual parcels or households.

**311 requests join failure.** At the time of writing, the 311 Requests dataset
has not successfully joined to the master due to neighbourhood key mismatches.
Domain 4 scores currently rely on trade permits only until this is resolved.

**Short-term rentals at ward level.** STR data is available at the electoral
ward level only. All neighbourhoods within a ward receive the same STR score,
which reduces sensitivity to within-ward variation.

---



---
title: "Winnipeg Neighbourhood Liveability Index"
subtitle: "Exploratory Factor Analysis · City of Winnipeg Open Data · 2021–2023"
author: "City of Winnipeg — Data & Analytics"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    theme: flatly
    highlight: tango
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: true
    toc_depth: 3
    number_sections: true
---
```{css, echo=FALSE}
body { font-family: 'Segoe UI', Arial, sans-serif; font-size: 14px; color: #2c3e50; }
h1, h2, h3 { color: #1a3a5c; }
.section-box {
  background: #f4f8fb; border-left: 4px solid #2980b9;
  padding: 12px 18px; margin-bottom: 18px; border-radius: 0 6px 6px 0;
}
.callout-warn {
  background: #fef9e7; border-left: 4px solid #f39c12;
  padding: 12px 18px; margin-bottom: 18px; border-radius: 0 6px 6px 0;
}
.callout-ok {
  background: #e9f7ef; border-left: 4px solid #27ae60;
  padding: 12px 18px; margin-bottom: 18px; border-radius: 0 6px 6px 0;
}
```
```{r setup2, include=FALSE}
knitr::opts_chunk$set(echo=FALSE, warning=FALSE, message=FALSE,
                      fig.width=10, fig.height=6, dpi=150)

library(dplyr); library(tidyr); library(tibble); library(stringr)
library(lubridate); library(purrr); library(scales)
library(psych); library(GPArotation)
library(DT); library(ggplot2); library(forcats)

INDEX_START  <- as.Date("2021-01-01")
INDEX_END    <- as.Date("2023-12-31")
NA_THRESHOLD <- 20

indicator_cols <- c(
  "total_living_area", "year_built", "rooms",
  "parks_count", "parks_total_area_ha", "tree_count",
  "substance_incidents_total", "naloxone_incidents_total",
  "naloxone_doses_total", "wfps_incidents_total",
  "wfps_median_response_mins", "trade_permits_total",
  "bldg_permits_total", "bldg_value_total", "bldg_major_projects",
  "rooming_complaints_total", "rooming_proactive_total",
  "population_total", "population_in_mbm", "land_area_in_square_km"
)
```

---

::: {.section-box}
This report derives the **Winnipeg Neighbourhood Liveability Index (WNLI)** using
Exploratory Factor Analysis (EFA) applied to parcel-level assessment data enriched
with neighbourhood-level indicators from City of Winnipeg open datasets.
The index aggregates five thematic domains: **Housing**, **Green Amenity**,
**Safety & Disorder**, **Built Environment Activity**, and **Social Vulnerability**.
All time-sensitive indicators are restricted to the **2021–2023** reference window.
:::

---

# Data & Indicator Selection

## Source Datasets

The following datasets were joined onto the assessment parcel anchor
(`master_merged`, 245,130 parcels × 110 columns) using `neighbourhood_area`
as the common key:
```{r source-table}
sources <- tibble::tibble(
  Dataset               = c("Assessment Parcels", "Parks & Open Space",
                             "Tree Inventory", "Substance Use Incidents",
                             "Naloxone Administrations", "WFPS Call Logs",
                             "Trade Permits", "Building Permits",
                             "Rooming House Enforcement", "Higher Poverty Areas"),
  Join_Key              = c("— (anchor)", rep("neighbourhood_area", 9)),
  Date_Filter           = c("Full dataset", "Full dataset", "Full dataset",
                             "2021–2023", "2021–2023", "2021–2023",
                             "2021–2023", "2021–2023", "2021–2023",
                             "2021 Census only"),
  Thematic_Domain       = c("Housing", "Green Amenity", "Green Amenity",
                             "Safety & Disorder", "Safety & Disorder",
                             "Safety & Disorder", "Built Activity",
                             "Built Activity", "Social Vulnerability",
                             "Social Vulnerability")
)

DT::datatable(
  sources,
  rownames  = FALSE,
  caption   = "Table 1 — Source datasets and join configuration",
  options   = list(pageLength = 10, dom = "ftp",
                   columnDefs = list(list(className = "dt-left", targets = "_all"))),
  class     = "stripe hover compact"
)
```

## Candidate Indicators

Twenty neighbourhood-level indicators were nominated across five thematic domains
prior to the missing-value screening described in the next section.
```{r indicator-table}
indicators <- tibble::tibble(
  Indicator                  = indicator_cols,
  Domain                     = c(
    "Housing", "Housing", "Housing",
    "Green Amenity", "Green Amenity", "Green Amenity",
    "Safety & Disorder", "Safety & Disorder", "Safety & Disorder",
    "Safety & Disorder", "Safety & Disorder",
    "Built Activity", "Built Activity", "Built Activity", "Built Activity",
    "Social Vulnerability", "Social Vulnerability",
    "Social Vulnerability", "Social Vulnerability", "Social Vulnerability"
  ),
  Description = c(
    "Mean total living area (sq ft)", "Mean year built", "Mean room count",
    "Number of parks within neighbourhood", "Total park area (hectares)",
    "Street tree count",
    "Substance-related EMS incidents", "Naloxone deployment incidents",
    "Naloxone doses administered", "WFPS total incidents",
    "WFPS median response time (mins)",
    "Total trade permits issued", "Total building permits",
    "Total declared construction value ($)", "Major construction projects count",
    "Rooming house complaint-driven inspections",
    "Rooming house proactive enforcement actions",
    "Total population (2021 Census)", "Population below MBM poverty line",
    "Land area (sq km)"
  )
)

DT::datatable(
  indicators,
  rownames  = FALSE,
  filter    = "top",
  caption   = "Table 2 — Candidate indicators (searchable by domain or name)",
  options   = list(pageLength = 10, dom = "ftp",
                   columnDefs = list(list(className = "dt-left", targets = "_all"))),
  class     = "stripe hover compact"
)
```

---

# Missing Value Audit {.tabset}

## Methodology

::: {.callout-warn}
**Decision rule:** Any indicator column where more than **`r NA_THRESHOLD`%** of
rows are missing is flagged **REMOVE** and excluded from EFA. Columns between
5–20% missing are flagged **REVIEW** and retained with median imputation.
Columns below 5% missing are flagged **OK**.
:::

The audit is computed across all 245,130 parcel rows in `master_merged`.
Columns with high sparsity typically arise from neighbourhood-level lookups
that could not be matched (e.g., newer subdivisions absent from the poverty
areas census table, or rooming house data restricted to inner-city neighbourhoods).

## Audit Results
```{r na-audit}
na_audit <- master_merged |>
  dplyr::select(dplyr::all_of(indicator_cols)) |>
  dplyr::summarise(dplyr::across(
    dplyr::everything(),
    \(x) round(mean(is.na(x)) * 100, 1)
  )) |>
  tidyr::pivot_longer(dplyr::everything(),
                      names_to  = "Indicator",
                      values_to = "Pct_NA") |>
  dplyr::mutate(
    Domain = dplyr::case_when(
      Indicator %in% c("total_living_area","year_built","rooms")
        ~ "Housing",
      Indicator %in% c("parks_count","parks_total_area_ha","tree_count")
        ~ "Green Amenity",
      Indicator %in% c("substance_incidents_total","naloxone_incidents_total",
                       "naloxone_doses_total","wfps_incidents_total",
                       "wfps_median_response_mins")
        ~ "Safety & Disorder",
      Indicator %in% c("trade_permits_total","bldg_permits_total",
                       "bldg_value_total","bldg_major_projects")
        ~ "Built Activity",
      TRUE ~ "Social Vulnerability"
    ),
    Decision = dplyr::case_when(
      Pct_NA >  NA_THRESHOLD ~ "REMOVE",
      Pct_NA >  5            ~ "REVIEW",
      TRUE                   ~ "OK"
    )
  ) |>
  dplyr::arrange(dplyr::desc(Pct_NA))

cols_to_remove <- na_audit |>
  dplyr::filter(Decision == "REMOVE") |>
  dplyr::pull(Indicator)

cols_for_efa <- setdiff(indicator_cols, cols_to_remove)

DT::datatable(
  na_audit,
  rownames  = FALSE,
  filter    = "top",
  caption   = "Table 3 — Missing value audit (searchable; sort by % NA or Decision)",
  options   = list(
    pageLength = 20,
    dom        = "ftp",
    columnDefs = list(list(className = "dt-center", targets = c(1, 3)))
  ),
  class = "stripe hover compact"
) |>
  DT::formatStyle(
    "Decision",
    backgroundColor = DT::styleEqual(
      c("REMOVE", "REVIEW", "OK"),
      c("#fde8e8",  "#fef9e7", "#e9f7ef")
    ),
    color = DT::styleEqual(
      c("REMOVE", "REVIEW", "OK"),
      c("#c0392b",  "#d68910", "#1e8449")
    ),
    fontWeight = "bold"
  ) |>
  DT::formatStyle("Pct_NA",
    background = DT::styleColorBar(c(0, 100), "#d6eaf8"),
    backgroundSize = "100% 80%",
    backgroundRepeat = "no-repeat",
    backgroundPosition = "center"
  )
```

## Summary
```{r na-summary}
na_summary <- na_audit |>
  dplyr::count(Decision) |>
  dplyr::mutate(
    Pct_of_Indicators = round(n / nrow(na_audit) * 100, 1),
    Action = dplyr::case_when(
      Decision == "REMOVE" ~ "Excluded from EFA",
      Decision == "REVIEW" ~ "Retained — median imputed",
      TRUE                 ~ "Retained as-is"
    )
  ) |>
  dplyr::rename(Count = n)

DT::datatable(
  na_summary,
  rownames = FALSE,
  caption  = "Table 4 — Audit decision summary",
  options  = list(dom = "t", pageLength = 5,
                  columnDefs = list(list(className = "dt-center", targets = c(1,2)))),
  class    = "stripe compact"
) |>
  DT::formatStyle(
    "Decision",
    backgroundColor = DT::styleEqual(
      c("REMOVE", "REVIEW", "OK"),
      c("#fde8e8", "#fef9e7", "#e9f7ef")
    ),
    color = DT::styleEqual(
      c("REMOVE", "REVIEW", "OK"),
      c("#c0392b", "#d68910", "#1e8449")
    ),
    fontWeight = "bold"
  )
```

::: {.callout-warn}
**Removed indicators:** `r paste(cols_to_remove, collapse = ", ")`.
These columns exceed the 20% NA threshold — most reflect inner-city–only
administrative records (rooming house enforcement, poverty census linkage)
that do not generalise across all Winnipeg neighbourhoods.
:::

---

# Neighbourhood-Level Aggregation

::: {.section-box}
EFA operates at the **neighbourhood level** (one row per neighbourhood), not
the parcel level. All retained indicators are averaged up from parcels to their
`neighbourhood_area` before scaling. Neighbourhoods where more than 50% of
retained indicators remain missing after aggregation are dropped.
:::
```{r neigh-agg}
neigh_efa <- master_merged |>
  dplyr::group_by(neighbourhood_area) |>
  dplyr::summarise(
    dplyr::across(dplyr::all_of(cols_for_efa),
                  \(x) mean(x, na.rm = TRUE)),
    .groups = "drop"
  ) |>
  dplyr::filter(
    rowMeans(is.na(dplyr::across(dplyr::all_of(cols_for_efa)))) < 0.5
  )
```
```{r neigh-preview}
DT::datatable(
  neigh_efa |>
    dplyr::mutate(dplyr::across(where(is.numeric), \(x) round(x, 2))),
  rownames  = FALSE,
  filter    = "top",
  caption   = paste0("Table 5 — Neighbourhood-level indicator matrix (",
                     nrow(neigh_efa), " neighbourhoods × ",
                     length(cols_for_efa), " indicators). Searchable."),
  options   = list(
    pageLength  = 15,
    scrollX     = TRUE,
    dom         = "fltip",
    autoWidth   = TRUE
  ),
  class = "stripe hover compact"
)
```

---

# EFA Pre-Checks {.tabset}

## Scaling & Variance
```{r scaling}
efa_matrix <- neigh_efa |>
  dplyr::select(dplyr::all_of(cols_for_efa)) |>
  dplyr::mutate(dplyr::across(dplyr::everything(), as.numeric)) |>
  dplyr::mutate(dplyr::across(
    dplyr::everything(),
    \(x) ifelse(is.na(x), median(x, na.rm = TRUE), x)
  )) |>
  scale() |>
  as.data.frame()

nzv      <- sapply(efa_matrix, sd, na.rm = TRUE)
efa_matrix <- efa_matrix[, nzv > 0.01]
```

After z-score standardisation and median imputation of residual NAs,
**`r ncol(efa_matrix)`** indicators with non-zero variance enter the EFA.
Near-zero-variance columns (SD ≤ 0.01 post-scaling) are automatically
removed to prevent matrix singularity.

## KMO & Bartlett
```{r precheck}
cor_matrix  <- cor(efa_matrix, use = "pairwise.complete.obs")
kmo_result  <- psych::KMO(cor_matrix)
bart_result <- psych::cortest.bartlett(cor_matrix, n = nrow(efa_matrix))

precheck_tbl <- tibble::tibble(
  Test      = c("KMO — Overall MSA",
                "Bartlett's Test — Chi-squared",
                "Bartlett's Test — p-value"),
  Value     = c(round(kmo_result$MSA, 3),
                round(bart_result$chisq, 2),
                format.pval(bart_result$p.value, digits = 3)),
  Threshold = c("> 0.60 (acceptable), > 0.80 (meritorious)",
                "—",
                "< 0.05 (reject sphericity null)"),
  Result    = c(
    ifelse(kmo_result$MSA >= 0.8, "Meritorious",
           ifelse(kmo_result$MSA >= 0.6, "Acceptable", "Poor")),
    "—",
    ifelse(bart_result$p.value < 0.05, "PASS", "FAIL")
  )
)

DT::datatable(
  precheck_tbl,
  rownames = FALSE,
  caption  = "Table 6 — EFA suitability pre-checks",
  options  = list(dom = "t", pageLength = 5,
                  columnDefs = list(list(className = "dt-left", targets = "_all"))),
  class    = "stripe compact"
) |>
  DT::formatStyle(
    "Result",
    color      = DT::styleEqual(
      c("Meritorious","Acceptable","Poor","PASS","FAIL","—"),
      c("#1e8449","#d68910","#c0392b","#1e8449","#c0392b","#555")
    ),
    fontWeight = "bold"
  )
```

## Per-Variable MSA
```{r kmo-var}
kmo_var <- tibble::tibble(
  Indicator = names(kmo_result$MSAi),
  MSA       = round(kmo_result$MSAi, 3),
  Flag      = dplyr::case_when(
    kmo_result$MSAi >= 0.80 ~ "Meritorious",
    kmo_result$MSAi >= 0.70 ~ "Middling",
    kmo_result$MSAi >= 0.60 ~ "Mediocre",
    kmo_result$MSAi >= 0.50 ~ "Miserable",
    TRUE                    ~ "Unacceptable — consider dropping"
  )
) |> dplyr::arrange(MSA)

DT::datatable(
  kmo_var,
  rownames = FALSE,
  filter   = "top",
  caption  = "Table 7 — Per-variable KMO MSA values (sort ascending to find weak indicators)",
  options  = list(pageLength = 15, dom = "ftp",
                  columnDefs = list(list(className = "dt-center", targets = 1))),
  class    = "stripe hover compact"
) |>
  DT::formatStyle(
    "Flag",
    backgroundColor = DT::styleEqual(
      c("Meritorious","Middling","Mediocre","Miserable","Unacceptable — consider dropping"),
      c("#e9f7ef","#eafaf1","#fef9e7","#fdebd0","#fde8e8")
    ),
    color = DT::styleEqual(
      c("Meritorious","Middling","Mediocre","Miserable","Unacceptable — consider dropping"),
      c("#1e8449","#1e8449","#d68910","#ca6f1e","#c0392b")
    ),
    fontWeight = "bold"
  )
```

---

# Factor Extraction {.tabset}

## Parallel Analysis

::: {.section-box}
Parallel analysis compares observed eigenvalues against eigenvalues from
random correlation matrices of the same size. Factors are retained where
the observed eigenvalue exceeds the simulated 95th percentile.
The scree plot below assists in determining the optimal number of factors **`n_factors`**.
After inspecting the plot, update `n_factors` in the chunk below before
proceeding to the EFA solution.
:::
```{r parallel, fig.cap="Figure 1 — Parallel Analysis Scree Plot. Retain factors whose observed eigenvalue (blue) exceeds the simulated line (red)."}
psych::fa.parallel(
  efa_matrix,
  fm     = "ml",
  fa     = "fa",
  n.iter = 100,
  main   = "Parallel Analysis — Winnipeg Liveability Indicators"
)
```

## EFA Solution
```{r efa}
# ── Update n_factors based on parallel analysis output above ──
n_factors <- 4

efa_result <- psych::fa(
  r        = efa_matrix,
  nfactors = n_factors,
  rotate   = "oblimin",
  fm       = "ml",
  scores   = "regression"
)
```
```{r efa-variance}
vaccounted <- as.data.frame(efa_result$Vaccounted) |>
  tibble::rownames_to_column("Statistic") |>
  dplyr::mutate(dplyr::across(where(is.numeric), \(x) round(x, 3)))

DT::datatable(
  vaccounted,
  rownames = FALSE,
  caption  = "Table 8 — Variance accounted for by each factor",
  options  = list(dom = "t", pageLength = 10,
                  columnDefs = list(list(className = "dt-center", targets = -1))),
  class    = "stripe compact"
)
```

## Factor Loadings

::: {.section-box}
Loadings ≥ |0.30| are considered meaningful. Use this table to assign
substantive labels to each factor (edit `factor_labels` in the next section
accordingly). Indicators with no loading ≥ |0.30| on any factor may be
candidates for removal in a confirmatory model.
:::
```{r loadings-table}
loadings_df <- as.data.frame(unclass(efa_result$loadings)) |>
  tibble::rownames_to_column("Indicator") |>
  dplyr::mutate(dplyr::across(where(is.numeric), \(x) round(x, 3)))

DT::datatable(
  loadings_df,
  rownames = FALSE,
  filter   = "top",
  caption  = "Table 9 — EFA factor loadings (|≥ 0.30| are meaningful; searchable by indicator name)",
  options  = list(
    pageLength = 20,
    scrollX    = TRUE,
    dom        = "ftp",
    columnDefs = list(list(className = "dt-center", targets = -1))
  ),
  class = "stripe hover compact"
) |>
  DT::formatStyle(
    colnames(loadings_df)[-1],
    backgroundColor = DT::styleInterval(
      c(-0.30, 0.30),
      c("#fde8e8", "white", "#e9f7ef")
    )
  )
```

## Factor Intercorrelations
```{r phi}
phi_df <- as.data.frame(round(efa_result$Phi, 3)) |>
  tibble::rownames_to_column("Factor")

DT::datatable(
  phi_df,
  rownames = FALSE,
  caption  = "Table 10 — Factor intercorrelation matrix (oblimin rotation)",
  options  = list(dom = "t", pageLength = 10,
                  columnDefs = list(list(className = "dt-center", targets = -1))),
  class    = "stripe compact"
)
```

---

# Liveability Index Construction {.tabset}

## Factor Scores
```{r factor-scores}
# ── Assign factor labels after inspecting Table 9 loadings ──
factor_labels <- c(
  "F1_GreenAmenity",
  "F2_SafetyDisorder",
  "F3_BuiltActivity",
  "F4_SocialVulnerability"
)

factor_scores <- as.data.frame(efa_result$scores)
names(factor_scores) <- factor_labels[seq_len(n_factors)]

neigh_scores <- dplyr::bind_cols(
  neigh_efa |> dplyr::select(neighbourhood_area),
  factor_scores
)
```
```{r scores-table}
DT::datatable(
  neigh_scores |>
    dplyr::mutate(dplyr::across(where(is.numeric), \(x) round(x, 3))),
  rownames = FALSE,
  filter   = "top",
  caption  = "Table 11 — Neighbourhood factor scores (searchable by neighbourhood name)",
  options  = list(
    pageLength = 15,
    scrollX    = TRUE,
    dom        = "fltip"
  ),
  class = "stripe hover compact"
)
```

## Composite Index

::: {.section-box}
**Sign convention:** Higher scores always indicate *better* liveability.
Factors where high scores represent *worse* conditions (e.g., more disorder,
more poverty) are negated before averaging. The composite is then rescaled
to a **0–100** scale for interpretability.
:::
```{r composite}
neigh_scores <- neigh_scores |>
  dplyr::mutate(
    F2_SafetyDisorder_inv      = -F2_SafetyDisorder,
    F4_SocialVulnerability_inv = -F4_SocialVulnerability,
    liveability_raw   = rowMeans(dplyr::pick(
      F1_GreenAmenity,
      F2_SafetyDisorder_inv,
      F3_BuiltActivity,
      F4_SocialVulnerability_inv
    ), na.rm = TRUE),
    liveability_score = scales::rescale(liveability_raw, to = c(0, 100))
  ) |>
  dplyr::arrange(dplyr::desc(liveability_score))
```
```{r composite-table}
DT::datatable(
  neigh_scores |>
    dplyr::select(
      neighbourhood_area, liveability_score,
      F1_GreenAmenity, F2_SafetyDisorder, F3_BuiltActivity, F4_SocialVulnerability
    ) |>
    dplyr::mutate(dplyr::across(where(is.numeric), \(x) round(x, 2))),
  rownames = FALSE,
  filter   = "top",
  caption  = "Table 12 — Winnipeg Neighbourhood Liveability Index scores (searchable; sort by score)",
  options  = list(
    pageLength = 20,
    scrollX    = TRUE,
    dom        = "fltip",
    columnDefs = list(list(className = "dt-center", targets = -1))
  ),
  class = "stripe hover compact"
) |>
  DT::formatStyle(
    "liveability_score",
    background         = DT::styleColorBar(c(0, 100), "#aed6f1"),
    backgroundSize     = "100% 80%",
    backgroundRepeat   = "no-repeat",
    backgroundPosition = "center"
  ) |>
  DT::formatStyle(
    "liveability_score",
    color      = DT::styleInterval(c(33, 66), c("#c0392b","#d68910","#1e8449")),
    fontWeight = "bold"
  )
```

---

# Results Visualisation
```{r plot-top-bottom, fig.cap="Figure 2 — Top and bottom 20 Winnipeg neighbourhoods by Liveability Index score (0–100 scale)."}
top_bottom <- dplyr::bind_rows(
  head(neigh_scores, 20) |> dplyr::mutate(Group = "Top 20"),
  tail(neigh_scores, 20) |> dplyr::mutate(Group = "Bottom 20")
)

ggplot2::ggplot(
  top_bottom,
  ggplot2::aes(
    x    = liveability_score,
    y    = forcats::fct_reorder(neighbourhood_area, liveability_score),
    fill = Group
  )
) +
  ggplot2::geom_col(width = 0.7) +
  ggplot2::geom_text(
    ggplot2::aes(label = round(liveability_score, 1)),
    hjust = -0.15, size = 3, colour = "#2c3e50"
  ) +
  ggplot2::scale_fill_manual(
    values = c("Top 20" = "#27ae60", "Bottom 20" = "#e74c3c")
  ) +
  ggplot2::scale_x_continuous(limits = c(0, 105), expand = c(0, 0)) +
  ggplot2::labs(
    title    = "Winnipeg Neighbourhood Liveability Index",
    subtitle = paste0(n_factors, "-factor EFA composite · 0–100 scale · 2021–2023 reference period"),
    x        = "Liveability Score",
    y        = NULL,
    fill     = NULL,
    caption  = "Source: City of Winnipeg Open Data · Parcel-level assessment anchor · EFA via psych package"
  ) +
  ggplot2::theme_minimal(base_size = 11) +
  ggplot2::theme(
    legend.position  = "top",
    panel.grid.major.y = ggplot2::element_blank(),
    plot.title       = ggplot2::element_text(face = "bold", colour = "#1a3a5c"),
    plot.subtitle    = ggplot2::element_text(colour = "#555"),
    plot.caption     = ggplot2::element_text(colour = "#888", size = 9)
  )
```

---

# Export Outputs
```{r export}
readr::write_csv(neigh_scores,  "winnipeg_liveability_scores.csv")
readr::write_csv(na_audit,      "winnipeg_na_audit.csv")

loadings_export <- as.data.frame(unclass(efa_result$loadings)) |>
  tibble::rownames_to_column("indicator")
readr::write_csv(loadings_export, "winnipeg_efa_loadings.csv")
```

Three files are written to the working directory upon knitting:

| File | Contents |
|---|---|
| `winnipeg_liveability_scores.csv` | Neighbourhood scores + factor scores |
| `winnipeg_na_audit.csv` | Full missing value audit with flags |
| `winnipeg_efa_loadings.csv` | Factor loading matrix |

---


---

# Parcel-Level Liveability Dataset {.tabset}

## Overview

::: {.section-box}
The neighbourhood-level liveability index and its four factor scores are
joined back onto every assessment parcel. Each row represents one property,
searchable by address, neighbourhood, or any housing characteristic.
Factor score columns are colour-coded by quartile so spatial clustering of
liveability is immediately visible at the parcel level.
:::

## Build Parcel Dataset
```{r parcel-join, include=FALSE}
# ── Factor label lookup (mirrors factor_labels defined earlier) ──
factor_col_lookup <- tibble::tibble(
  domain   = c("Housing", "Green Amenity",
                "Safety & Disorder", "Built Activity", "Social Vulnerability"),
  factor   = c("F1_GreenAmenity", "F1_GreenAmenity",
                "F2_SafetyDisorder", "F3_BuiltActivity", "F4_SocialVulnerability")
)

# ── Indicators retained under each factor (from loadings inspection) ──
f1_vars <- c("parks_count", "parks_total_area_ha", "tree_count")
f2_vars <- c("substance_incidents_total", "naloxone_incidents_total",
              "naloxone_doses_total", "wfps_incidents_total",
              "wfps_median_response_mins")
f3_vars <- c("trade_permits_total", "bldg_permits_total",
              "bldg_value_total",   "bldg_major_projects")
f4_vars <- c("population_total", "population_in_mbm",
              "land_area_in_square_km")

# ── Core parcel / address / housing columns ──
parcel_core <- c(
  "roll_number", "full_address", "street_number", "street_name",
  "street_type", "neighbourhood_area", "market_region",
  "building_type", "total_living_area", "year_built", "rooms",
  "basement", "basement_finish", "air_conditioning",
  "fire_place", "attached_garage", "detached_garage", "pool",
  "assessed_land_area", "total_assessed_value",
  "total_proposed_assessment_value", "zoning",
  "property_use_code", "centroid_lat", "centroid_lon"
)

# ── Join liveability scores onto master_merged ──
parcel_liveability <- master_merged |>
  dplyr::left_join(
    neigh_scores |>
      dplyr::select(
        neighbourhood_area,
        liveability_score,
        F1_GreenAmenity,
        F2_SafetyDisorder,
        F3_BuiltActivity,
        F4_SocialVulnerability
      ),
    by = "neighbourhood_area"
  ) |>
  dplyr::select(
    dplyr::all_of(parcel_core),
    liveability_score,
    F1_GreenAmenity, F2_SafetyDisorder, F3_BuiltActivity, F4_SocialVulnerability,
    dplyr::all_of(intersect(c(f1_vars, f2_vars, f3_vars, f4_vars),
                             names(master_merged)))
  ) |>
  dplyr::mutate(
    total_assessed_value           = as.numeric(total_assessed_value),
    total_proposed_assessment_value = as.numeric(total_proposed_assessment_value),
    dplyr::across(where(is.numeric), \(x) round(x, 2))
  ) |>
  dplyr::arrange(dplyr::desc(liveability_score))

# ── Friendly column names for display ──
display_names <- c(
  "Roll Number", "Full Address", "Street No.", "Street Name", "Street Type",
  "Neighbourhood", "Market Region",
  "Building Type", "Living Area (sq ft)", "Year Built", "Rooms",
  "Basement", "Basement Finished", "Air Conditioning",
  "Fireplace", "Attached Garage", "Detached Garage", "Pool",
  "Land Area (sq ft)", "Assessed Value ($)", "Proposed Value ($)",
  "Zoning", "Property Use", "Latitude", "Longitude",
  "Liveability Score",
  "F1: Green Amenity", "F2: Safety & Disorder",
  "F3: Built Activity", "F4: Social Vulnerability",
  # factor indicator variables — keep original names trimmed
  intersect(c(f1_vars, f2_vars, f3_vars, f4_vars), names(master_merged))
)

# Align name vector length safely
names(parcel_liveability) <- display_names[seq_len(ncol(parcel_liveability))]
```

## Parcel Table

::: {.callout-warn}
**How to use this table:** Search any address, neighbourhood, or street name
using the global search box. Use column filters (top row) to narrow by
building type, year built, zoning, etc.
Factor columns are colour-coded: 🟢 **top quartile** · 🟡 **second quartile**
· 🟠 **third quartile** · 🔴 **bottom quartile**.
For Safety & Disorder, the colour scale is **inverted** (lower = better).
:::
```{r parcel-dt}
# ── Identify factor column positions (0-indexed for DT) ──
col_names_display <- names(parcel_liveability)

idx_score <- which(col_names_display == "Liveability Score")       - 1L
idx_f1    <- which(col_names_display == "F1: Green Amenity")       - 1L
idx_f2    <- which(col_names_display == "F2: Safety & Disorder")   - 1L
idx_f3    <- which(col_names_display == "F3: Built Activity")      - 1L
idx_f4    <- which(col_names_display == "F4: Social Vulnerability") - 1L

# ── Quartile colour palettes ──
# Green Amenity, Built Activity, Social Vulnerability: high = good = green
pos_pal <- c("#c0392b", "#e67e22", "#f1c40f", "#27ae60")   # Q1→Q4

# Safety & Disorder: high = bad = red (inverted)
neg_pal <- c("#27ae60", "#f1c40f", "#e67e22", "#c0392b")   # Q1→Q4

score_pal <- c("#e74c3c","#e67e22","#f7dc6f","#a9dfbf","#27ae60")

quartile_breaks <- function(x) quantile(x, probs = c(0.25, 0.50, 0.75), na.rm = TRUE)

q_score <- quantile(parcel_liveability[["Liveability Score"]],
                    probs = seq(0, 1, 0.2), na.rm = TRUE)
q_f1    <- quartile_breaks(parcel_liveability[["F1: Green Amenity"]])
q_f2    <- quartile_breaks(parcel_liveability[["F2: Safety & Disorder"]])
q_f3    <- quartile_breaks(parcel_liveability[["F3: Built Activity"]])
q_f4    <- quartile_breaks(parcel_liveability[["F4: Social Vulnerability"]])

DT::datatable(
  parcel_liveability,
  rownames   = FALSE,
  filter     = "top",
  caption    = paste0(
    "Table 13 — Parcel-level Liveability Dataset  (",
    format(nrow(parcel_liveability), big.mark = ","),
    " parcels). Search by address, neighbourhood, or any column filter."
  ),
  extensions = c("Buttons", "FixedColumns", "Scroller"),
  options    = list(
    pageLength   = 25,
    scrollX      = TRUE,
    scrollY      = "520px",
    scroller     = TRUE,
    deferRender  = TRUE,
    dom          = "Bfltip",
    fixedColumns = list(leftColumns = 2L),   # freeze Roll Number + Address
    buttons      = list(
      list(extend = "csv",   filename = "winnipeg_parcel_liveability"),
      list(extend = "excel", filename = "winnipeg_parcel_liveability"),
      "colvis"               # toggle column visibility
    ),
    columnDefs   = list(
      list(className = "dt-center",
           targets   = c(idx_score, idx_f1, idx_f2, idx_f3, idx_f4)),
      list(visible = FALSE,
           targets  = c(
             which(col_names_display %in%
                     c("Street No.", "Street Name", "Street Type",
                       "Latitude", "Longitude", "Roll Number")) - 1L
           ))
    )
  ),
  class = "stripe hover compact nowrap"
) |>

  # ── Liveability Score — continuous blue bar + coloured text ──
  DT::formatStyle(
    "Liveability Score",
    background         = DT::styleColorBar(c(0, 100), "#aed6f1"),
    backgroundSize     = "100% 75%",
    backgroundRepeat   = "no-repeat",
    backgroundPosition = "center",
    color              = DT::styleInterval(
      c(20, 40, 60, 80),
      c("#922b21","#ca6f1e","#7d6608","#1a5276","#1e8449")
    ),
    fontWeight = "bold"
  ) |>

  # ── F1: Green Amenity — quartile green ramp (high = good) ──
  DT::formatStyle(
    "F1: Green Amenity",
    backgroundColor = DT::styleInterval(q_f1, c(pos_pal[1], pos_pal[2],
                                                  pos_pal[3], pos_pal[4])),
    color           = "white",
    fontWeight      = "bold"
  ) |>

  # ── F2: Safety & Disorder — inverted (high = bad = red) ──
  DT::formatStyle(
    "F2: Safety & Disorder",
    backgroundColor = DT::styleInterval(q_f2, c(neg_pal[1], neg_pal[2],
                                                  neg_pal[3], neg_pal[4])),
    color           = "white",
    fontWeight      = "bold"
  ) |>

  # ── F3: Built Activity — quartile green ramp ──
  DT::formatStyle(
    "F3: Built Activity",
    backgroundColor = DT::styleInterval(q_f3, c(pos_pal[1], pos_pal[2],
                                                  pos_pal[3], pos_pal[4])),
    color           = "white",
    fontWeight      = "bold"
  ) |>

  # ── F4: Social Vulnerability — inverted (high vulnerability = red) ──
  DT::formatStyle(
    "F4: Social Vulnerability",
    backgroundColor = DT::styleInterval(q_f4, c(neg_pal[1], neg_pal[2],
                                                  neg_pal[3], neg_pal[4])),
    color           = "white",
    fontWeight      = "bold"
  ) |>

  # ── Assessed Value — formatted as dollar ──
  DT::formatCurrency(
    columns  = "Assessed Value ($)",
    currency = "$",
    digits   = 0
  ) |>
  DT::formatCurrency(
    columns  = "Proposed Value ($)",
    currency = "$",
    digits   = 0
  ) |>

  # ── Soft row stripe by building type ──
  DT::formatStyle(
    "Building Type",
    backgroundColor = DT::styleEqual(
      c("ONE STOREY","TWO STOREY","BI-LEVEL","SPLIT LEVEL","THREE STOREY"),
      c("#fdfefe",  "#f4f6f7",   "#eaf2ff",  "#fff9f0",    "#f9f0ff")
    )
  )
```

## Download & Notes

::: {.section-box}
**Download options** are available via the **CSV** and **Excel** buttons in the
table toolbar above. The **Col visibility** button lets you show or hide any column.

**Column groups by factor:**

| Factor | Colour Scale | Indicator Columns Included |
|---|---|---|
| F1: Green Amenity | 🟥→🟩 (high = good) | `parks_count`, `parks_total_area_ha`, `tree_count` |
| F2: Safety & Disorder | 🟩→🟥 (high = bad) | `substance_incidents_total`, `naloxone_incidents_total`, `naloxone_doses_total`, `wfps_incidents_total`, `wfps_median_response_mins` |
| F3: Built Activity | 🟥→🟩 (high = good) | `trade_permits_total`, `bldg_permits_total`, `bldg_value_total`, `bldg_major_projects` |
| F4: Social Vulnerability | 🟩→🟥 (high = bad) | `population_total`, `population_in_mbm`, `land_area_in_square_km` |

**Liveability Score** (0–100) is an equal-weighted composite of all four
inverted/signed factor scores, rescaled linearly.
The score is assigned at the **neighbourhood level** and is identical for all
parcels within the same neighbourhood.
:::

# Session Info
```{r session}
sessionInfo()
```

