---
title: "Winnipeg Open Datathon — Import & Diagnostics"
author: "Myo Minn Oo & Yong Jin"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    highlight: tango
    df_print: default
    code_folding: hide
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE
)
```

---

# Overview

This document covers the full pipeline for importing, caching, and diagnosing
11 datasets from the [Winnipeg Open Data Portal](https://data.winnipeg.ca).
The datasets span property assessment, public health, transit, land use, and
census data. The ultimate goal is to merge them into a parcel-level master
dataset anchored on **Assessment Parcels**.

---

# 1. API Configuration

Each dataset is referenced by its Socrata resource ID. Per-dataset row limits
are defined separately so they can be adjusted without touching the fetch logic.
`NULL` means no limit — `RSocrata::read.socrata()` will auto-paginate and
return all rows.

Note: Two datasets — **Transit Pass-ups** and **Short Term Rentals** — have no
neighbourhood identifier and will require a manual spatial join later.

```{r api-config}
api_list <- list(
  assessment_parcels        = "https://data.winnipeg.ca/resource/d4mq-wa44.json",
  parks_open_space          = "https://data.winnipeg.ca/resource/tx3d-pfxq.json",
  substance_use             = "https://data.winnipeg.ca/resource/6x82-bz5y.json",
  naloxone_administrations  = "https://data.winnipeg.ca/resource/qd6b-q49i.json",
  transit_passups           = "https://data.winnipeg.ca/resource/mer2-irmb.json",  # no neighbourhood
  census_households         = "https://data.winnipeg.ca/resource/nmk5-uwfw.json",
  requests_311              = "https://data.winnipeg.ca/resource/u7f6-5326.json",
  higher_poverty_areas      = "https://data.winnipeg.ca/resource/gg4w-peq6.json",
  trade_permits             = "https://data.winnipeg.ca/resource/urbd-qygv.json",
  rooming_house_enforcement = "https://data.winnipeg.ca/resource/vk2f-xwp7.json",
  short_term_rentals        = "https://data.winnipeg.ca/resource/74hr-f8ai.json"   # no neighbourhood
)

api_limits <- list(
  assessment_parcels        = 50000,
  parks_open_space          = NULL,
  substance_use             = NULL,
  naloxone_administrations  = NULL,
  transit_passups           = 50000,
  census_households         = NULL,
  requests_311              = 50000,
  higher_poverty_areas      = NULL,
  trade_permits             = 50000,
  rooming_house_enforcement = NULL,
  short_term_rentals        = 5000
)
```

---

# 2. Fetch Functions

With Claude's help, we have created three modular functions handle the import pipeline:

- `build_url()` — appends `?$limit=N` to a URL when a limit is specified
- `fetch_socrata()` — fetches a single dataset with `tryCatch` error handling;
  returns `NULL` on failure so the rest of the pipeline continues uninterrupted
- `fetch_all_socrata()` — orchestrates the full fetch using `purrr::map2()` to
  pair each URL with its limit, then reports any failures
  
  
```{r fetch-functions}
# Module 1: Build URL with optional row limit ----
build_url <- function(url, limit = NULL) {
  if (!base::is.null(limit)) {
    base::paste0(url, "?$limit=", limit)
  } else {
    url
  }
}

# Module 2: Fetch a single dataset with error handling ----
fetch_socrata <- function(url, limit = NULL) {
  final_url <- build_url(url, limit)
  base::message("Fetching: ", final_url)
  tryCatch(
    expr    = RSocrata::read.socrata(final_url),
    error   = \(e) {
      base::message("ERROR fetching ", final_url, "\n  → ", base::conditionMessage(e))
      NULL
    },
    warning = \(w) {
      base::message("WARNING fetching ", final_url, "\n  → ", base::conditionMessage(w))
      RSocrata::read.socrata(final_url)
    }
  )
}

# Module 3: Report failed datasets ----
report_failures <- function(data_list) {
  failed <- base::names(purrr::keep(data_list, base::is.null))
  if (base::length(failed) > 0) {
    base::message("\nFailed to load: ", base::paste(failed, collapse = ", "))
  } else {
    base::message("\nAll datasets loaded successfully.")
  }
  base::invisible(failed)
}

# Module 4: Fetch all datasets with per-dataset limits ----
fetch_all_socrata <- function(api_list, api_limits = NULL) {
  if (base::is.null(api_limits)) {
    data_list <- purrr::map(api_list, fetch_socrata)
  } else {
    data_list <- purrr::map2(
      api_list,
      api_limits,
      \(url, lim) fetch_socrata(url, limit = lim)
    )
  }
  report_failures(data_list)
  data_list
}
```

---

# 3. Import Data

Data is loaded from a local cache (`data/data_raw_list.RData`) if it exists,
avoiding unnecessary repeat API calls. If no cache is found, data is fetched
fresh from the API and saved for future sessions.
```{r import-data}
cache_path <- "data/data_raw_list.RData"

if (base::file.exists(cache_path)) {
  base::message("Loading from cache: ", cache_path)
  base::load(cache_path)
} else {
  base::message("Cache not found — fetching from API...")
  data_raw_list <- fetch_all_socrata(api_list, api_limits)
  base::save(data_raw_list, file = cache_path)
}
```

---

# 4. Structure Overview

Quick check of rows and columns per dataset to verify the pull was complete.
```{r structure-overview}
dim_summary <- purrr::imap_dfr(data_raw_list, \(df, name) {
  tibble::tibble(
    dataset = name,
    rows    = base::nrow(df),
    cols    = base::ncol(df),
    loaded  = !base::is.null(df)
  )
})

dim_summary |> 
  make_dt()
```

---

# 5. Join Key Scan

Scan all datasets for columns likely to serve as merge keys — neighbourhood,
ward, and ID fields.
```{r join-key-scan}
join_keys <- purrr::imap_dfr(data_raw_list, \(df, name) {
  if (base::is.null(df)) return(NULL)
  matched <- base::grep(
    pattern     = "neighbourhood|neigh|_id|ward",
    x           = base::names(df),
    value       = TRUE,
    ignore.case = TRUE
  )
  if (base::length(matched) == 0) return(NULL)
  tibble::tibble(
    dataset   = name,
    join_keys = base::paste(matched, collapse = ", ")
  )
})

join_keys |> 
  make_dt()
```

---

# 6. Column Diagnostics

The `check_all()` function runs a comprehensive column-level audit across every
dataset. For each column it reports:

- **n_total / n_valid / n_missing** — observation counts
- **pct_missing** — percentage of missing or blank values
- **min / max / mean / median / sd** — for numeric columns
- **min / max** (date range) — for date columns
- **n_unique / categories** — top categories with counts for character columns

---

## 6.1 Missing Values Summary

Columns with any missing values, ranked by severity.
```{r missing-values}
missing_summary <- diagnostics |>
  dplyr::filter(pct_missing > 0) |>
  dplyr::select(dataset, column, type, n_total, n_missing, pct_missing) |>
  dplyr::arrange(dplyr::desc(pct_missing)) 

missing_summary |> 
  make_dt()
```
```{r missing-flag}
# Columns with > 50% missing — candidates for exclusion ----
missing_summary |>
  dplyr::filter(pct_missing > 50) |>
  dplyr::select(dataset, column, pct_missing) |> 
  make_dt()
```

---

## 6.2 Numeric Column Statistics

Summary statistics for all numeric and integer columns across every dataset.
```{r numeric-stats}
diagnostics |>
  dplyr::filter(type %in% c("numeric", "integer")) |>
  dplyr::select(dataset, column, min, max, mean, median, sd, n_missing, pct_missing) |> 
  make_dt()
```

---

## 6.3 Categorical Column Valid Values

Valid categories and their frequencies for all character columns. Columns with
more than 10 unique values show the top 10 plus a count of remaining categories.
```{r categorical-values}
diagnostics |>
  dplyr::filter(type == "character", !base::is.na(categories)) |>
  dplyr::select(dataset, column, n_unique, categories) |> 
  make_dt()
```

---

## 6.4 Date Column Ranges

Date range (min → max) for all date and datetime columns — useful for
understanding temporal coverage before merging.
```{r date-ranges}
diagnostics |>
  dplyr::filter(type %in% c("POSIXct", "POSIXlt", "Date")) |>
  dplyr::select(dataset, column, min, max, n_unique) |> 
  make_dt()
```


---

# 8. Session Info
```{r session-info}
utils::sessionInfo()
```