---
title: "Winnipeg Open Datathon — Import & Diagnostics"
author: "Myo Minn Oo & Yong Jin"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    highlight: tango
    df_print: default
    code_folding: hide
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE
)
```

---

# Section 1: Data Import, Cleaning, And Processing

This document covers the full pipeline for importing, caching, and diagnosing
11 datasets from the [Winnipeg Open Data Portal](https://data.winnipeg.ca).
The datasets span property assessment, public health, transit, land use, and
census data. The ultimate goal is to merge them into a parcel-level master
dataset anchored on **Assessment Parcels**.

---

# 1. API Configuration

Each dataset is referenced by its Socrata resource ID. Per-dataset row limits
are defined separately so they can be adjusted without touching the fetch logic.
`NULL` means no limit — `RSocrata::read.socrata()` will auto-paginate and
return all rows.

Note: Two datasets — **Transit Pass-ups** and **Short Term Rentals** — have no
neighbourhood identifier and will require a manual spatial join later.

```{r api-config}
api_list <- list(
  assessment_parcels        = "https://data.winnipeg.ca/resource/d4mq-wa44.json",
  parks_open_space          = "https://data.winnipeg.ca/resource/tx3d-pfxq.json",
  substance_use             = "https://data.winnipeg.ca/resource/6x82-bz5y.json",
  naloxone_administrations  = "https://data.winnipeg.ca/resource/qd6b-q49i.json",
  transit_passups           = "https://data.winnipeg.ca/resource/mer2-irmb.json",  # no neighbourhood
  census_households         = "https://data.winnipeg.ca/resource/nmk5-uwfw.json",
  requests_311              = "https://data.winnipeg.ca/resource/u7f6-5326.json",
  higher_poverty_areas      = "https://data.winnipeg.ca/resource/gg4w-peq6.json",
  trade_permits             = "https://data.winnipeg.ca/resource/urbd-qygv.json",
  rooming_house_enforcement = "https://data.winnipeg.ca/resource/vk2f-xwp7.json",
  short_term_rentals        = "https://data.winnipeg.ca/resource/74hr-f8ai.json"   # no neighbourhood
)

api_limits <- list(
  assessment_parcels        = 50000,
  parks_open_space          = NULL,
  substance_use             = NULL,
  naloxone_administrations  = NULL,
  transit_passups           = 50000,
  census_households         = NULL,
  requests_311              = 50000,
  higher_poverty_areas      = NULL,
  trade_permits             = 50000,
  rooming_house_enforcement = NULL,
  short_term_rentals        = 5000
)
```

---

# 2. Fetch Functions

With Claude's help, we have created three modular functions handle the import pipeline:

- `build_url()` — appends `?$limit=N` to a URL when a limit is specified
- `fetch_socrata()` — fetches a single dataset with `tryCatch` error handling;
  returns `NULL` on failure so the rest of the pipeline continues uninterrupted
- `fetch_all_socrata()` — orchestrates the full fetch using `purrr::map2()` to
  pair each URL with its limit, then reports any failures
  
  
```{r fetch-functions}
# Module 1: Build URL with optional row limit ----
build_url <- function(url, limit = NULL) {
  if (!base::is.null(limit)) {
    base::paste0(url, "?$limit=", limit)
  } else {
    url
  }
}

# Module 2: Fetch a single dataset with error handling ----
fetch_socrata <- function(url, limit = NULL) {
  final_url <- build_url(url, limit)
  base::message("Fetching: ", final_url)
  tryCatch(
    expr    = RSocrata::read.socrata(final_url),
    error   = \(e) {
      base::message("ERROR fetching ", final_url, "\n  → ", base::conditionMessage(e))
      NULL
    },
    warning = \(w) {
      base::message("WARNING fetching ", final_url, "\n  → ", base::conditionMessage(w))
      RSocrata::read.socrata(final_url)
    }
  )
}

# Module 3: Report failed datasets ----
report_failures <- function(data_list) {
  failed <- base::names(purrr::keep(data_list, base::is.null))
  if (base::length(failed) > 0) {
    base::message("\nFailed to load: ", base::paste(failed, collapse = ", "))
  } else {
    base::message("\nAll datasets loaded successfully.")
  }
  base::invisible(failed)
}

# Module 4: Fetch all datasets with per-dataset limits ----
fetch_all_socrata <- function(api_list, api_limits = NULL) {
  if (base::is.null(api_limits)) {
    data_list <- purrr::map(api_list, fetch_socrata)
  } else {
    data_list <- purrr::map2(
      api_list,
      api_limits,
      \(url, lim) fetch_socrata(url, limit = lim)
    )
  }
  report_failures(data_list)
  data_list
}
```

---

# 3. Import Data

Data is loaded from a local cache (`data/data_raw_list.RData`) if it exists,
avoiding unnecessary repeat API calls. If no cache is found, data is fetched
fresh from the API and saved for future sessions.
```{r import-data}
cache_path <- "data/data_raw_list.RData"

if (base::file.exists(cache_path)) {
  base::message("Loading from cache: ", cache_path)
  base::load(cache_path)
} else {
  base::message("Cache not found — fetching from API...")
  data_raw_list <- fetch_all_socrata(api_list, api_limits)
  base::save(data_raw_list, file = cache_path)
}
```

---

# 4. Structure Overview

Quick check of rows and columns per dataset to verify the pull was complete.
```{r structure-overview}
dim_summary <- purrr::imap_dfr(data_raw_list, \(df, name) {
  tibble::tibble(
    dataset = name,
    rows    = base::nrow(df),
    cols    = base::ncol(df),
    loaded  = !base::is.null(df)
  )
})

dim_summary |> 
  make_dt()
```

---

# 5. Join Key Scan

Scan all datasets for columns likely to serve as merge keys — neighbourhood,
ward, and ID fields.
```{r join-key-scan}
join_keys <- purrr::imap_dfr(data_raw_list, \(df, name) {
  if (base::is.null(df)) return(NULL)
  matched <- base::grep(
    pattern     = "neighbourhood|neigh|_id|ward",
    x           = base::names(df),
    value       = TRUE,
    ignore.case = TRUE
  )
  if (base::length(matched) == 0) return(NULL)
  tibble::tibble(
    dataset   = name,
    join_keys = base::paste(matched, collapse = ", ")
  )
})

join_keys |> 
  make_dt()
```

---

# 6. Column Diagnostics

The `check_all()` function runs a comprehensive column-level audit across every
dataset. For each column it reports:

- **n_total / n_valid / n_missing** — observation counts
- **pct_missing** — percentage of missing or blank values
- **min / max / mean / median / sd** — for numeric columns
- **min / max** (date range) — for date columns
- **n_unique / categories** — top categories with counts for character columns

---

## 6.1 Missing Values Summary

Columns with any missing values, ranked by severity.
```{r missing-values}
missing_summary <- diagnostics |>
  dplyr::filter(pct_missing > 0) |>
  dplyr::select(dataset, column, type, n_total, n_missing, pct_missing) |>
  dplyr::arrange(dplyr::desc(pct_missing)) 

missing_summary |> 
  make_dt()
```
```{r missing-flag}
# Columns with > 50% missing — candidates for exclusion ----
missing_summary |>
  dplyr::filter(pct_missing > 50) |>
  dplyr::select(dataset, column, pct_missing) |> 
  make_dt()
```

---

## 6.2 Numeric Column Statistics

Summary statistics for all numeric and integer columns across every dataset.
```{r numeric-stats}
diagnostics |>
  dplyr::filter(type %in% c("numeric", "integer")) |>
  dplyr::select(dataset, column, min, max, mean, median, sd, n_missing, pct_missing) |> 
  make_dt()
```

---

## 6.3 Categorical Column Valid Values

Valid categories and their frequencies for all character columns. Columns with
more than 10 unique values show the top 10 plus a count of remaining categories.
```{r categorical-values}
diagnostics |>
  dplyr::filter(type == "character", !base::is.na(categories)) |>
  dplyr::select(dataset, column, n_unique, categories) |> 
  make_dt()
```

---

## 6.4 Date Column Ranges

Date range (min → max) for all date and datetime columns — useful for
understanding temporal coverage before merging.
```{r date-ranges}
diagnostics |>
  dplyr::filter(type %in% c("POSIXct", "POSIXlt", "Date")) |>
  dplyr::select(dataset, column, min, max, n_unique) |> 
  make_dt()
```


---

# 8. Session Info
```{r session-info}
utils::sessionInfo()
```



md---
title: "Winnipeg Neighbourhood Liveability Index — Methodology"
author: ""
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    highlight: tango
    df_print: default
---

---

# Section 2: Factor Analysis

The Winnipeg Neighbourhood Liveability Index (WNLI) is a composite score that
summarises the overall quality of life across Winnipeg neighbourhoods using
publicly available open data from the
[Winnipeg Open Data Portal](https://data.winnipeg.ca). The index aggregates
11 datasets spanning property assessment, public health, green space, municipal
services, housing vulnerability, and short-term rentals into a single 0–100
score per neighbourhood.

The index is designed to be **transparent**, **reproducible**, and
**comparable across neighbourhoods** — three properties that guided every
methodological decision documented here.

---

# 1. Common Approaches to Liveability Indices

There are three established methods for constructing composite liveability
indices. Each involves tradeoffs between simplicity, statistical rigour, and
communicability.

---

## 1.1 Equal-Weight Composite Scoring

The simplest approach. Each indicator is normalized to a common scale
(typically 0–100 or a z-score), then all indicators are averaged with equal
weight to produce a final score.

**Used by:** The Economist Intelligence Unit Global Liveability Index,
MERCER Quality of Living Survey.

**Strengths:**

- Fully transparent — every indicator contributes the same amount
- Easy to explain to non-technical audiences
- Simple to update as new data becomes available

**Weaknesses:**

- Implicitly assumes all dimensions of liveability are equally important
- Sensitive to the number of indicators per domain — domains with more
  indicators get more influence
- Does not account for statistical redundancy between correlated indicators

---

## 1.2 Domain-Weighted Composite Scoring

Indicators are grouped into thematic **domains** (e.g. safety, housing, green
space). Scores are computed within each domain, then domains are combined with
explicit weights — either equal across domains, or set by expert judgment or
community consultation.

**Used by:** The Canadian Index of Wellbeing, the UN Human Development Index,
most municipal and regional liveability indices.

**Strengths:**

- Balances domains regardless of how many indicators each contains
- Weights can be adjusted to reflect local priorities
- Results are interpretable at the domain level, not just the composite

**Weaknesses:**

- Domain weights are subjective unless derived from community input or research
- Requires clear theoretical justification for what belongs in each domain

> **This is the approach adopted for the WNLI.** Equal weights are applied
> across six domains in the first iteration. Sensitivity analysis will test
> alternative weighting schemes.

---

## 1.3 Principal Component Analysis / Factor Analysis

A data-driven approach where the statistical structure of the indicators
determines how they are grouped and weighted. PCA identifies linear combinations
of indicators that explain the most variance, while factor analysis uncovers
latent constructs underlying the observed variables.

**Used by:** Academic liveability research, neighbourhood deprivation indices,
Statistics Canada's Material and Social Deprivation Index.

**Strengths:**

- Weights are derived from the data, not subjective judgment
- Automatically handles correlated indicators by consolidating them
- Defensible in peer-reviewed or technical contexts

**Weaknesses:**

- Results are difficult to communicate to general audiences
- Component loadings can be unstable with small samples or high missingness
- Less practical for iterative updates as new data arrives

---

# 2. Normalization Methods

Before indicators can be combined into a composite score, they must be placed
on a common scale. The three most common normalization methods are:

| Method | Formula | Range | Best for |
|---|---|---|---|
| **Min-max** | (x − min) / (max − min) | 0 to 1 | Interpretable scores bounded between 0 and 100 |
| **Z-score** | (x − mean) / sd | Unbounded | Comparing deviations from the city average |
| **Rank percentile** | rank(x) / n | 0 to 1 | Robust to outliers; less sensitive to extreme values |

The WNLI uses **min-max normalization** scaled to 0–100. This produces scores
that are immediately interpretable — a neighbourhood scoring 75 performs better
than 75% of the possible range on that indicator. Z-scores will be computed
as a robustness check.

---

# 3. Domain Structure

The WNLI organises its indicators into **six domains**, each reflecting a
distinct dimension of neighbourhood liveability. The domain structure is
informed by the Canadian Index of Wellbeing framework and adapted to the
variables available in Winnipeg's open data.

---

## Domain 1 — Housing & Property

Reflects the quality, size, age, and value of the residential housing stock
in each neighbourhood.

| Indicator | Source | Direction |
|---|---|---|
| Median assessed property value | Assessment Parcels | ↑ Higher = better |
| Median living area (sq ft) | Assessment Parcels | ↑ Higher = better |
| Median housing age | Assessment Parcels | ↓ Older = worse |
| Total dwelling units | Assessment Parcels | ↑ Higher = better |
| Assessment value growth rate | Assessment Parcels | ↑ Higher = better |
| % residential land use | Assessment Parcels | ↑ Higher = better |

---

## Domain 2 — Green Space & Environment

Reflects access to parks and outdoor amenity within each neighbourhood.

| Indicator | Source | Direction |
|---|---|---|
| Parks per 1,000 residents | Parks & Open Space | ↑ Higher = better |
| Park area (ha) per 1,000 residents | Parks & Open Space | ↑ Higher = better |
| Active park & open space area (ha) | Parks & Open Space | ↑ Higher = better |
| Water feature area (ha) | Parks & Open Space | ↑ Higher = better |

---

## Domain 3 — Public Safety & Health

Reflects the burden of substance use-related incidents and opioid overdoses
within each neighbourhood during the 2021–2023 index window.

| Indicator | Source | Direction |
|---|---|---|
| Substance use incidents per 1,000 residents | Substance Use | ↓ Higher = worse |
| Opioid incidents per 1,000 residents | Substance Use | ↓ Higher = worse |
| Crystal meth incidents per 1,000 residents | Substance Use | ↓ Higher = worse |
| Naloxone administrations per 1,000 residents | Naloxone Administrations | ↓ Higher = worse |
| Average naloxone doses per incident | Naloxone Administrations | ↓ Higher = worse |

---

## Domain 4 — Municipal Services & Activity

Reflects both the demand for municipal services and the level of economic
activity and investment in each neighbourhood.

| Indicator | Source | Direction |
|---|---|---|
| 311 requests per 1,000 residents | 311 Requests | Ambiguous — see note |
| 311 request resolution rate (closed/total) | 311 Requests | ↑ Higher = better |
| Trade permits per 1,000 residents | Trade Permits | ↑ Higher = better |
| Trade permit issuance rate (issued/total) | Trade Permits | ↑ Higher = better |

> **Note on 311 requests:** A high volume of 311 requests can indicate either
> an engaged, civically active neighbourhood or a distressed neighbourhood with
> high service needs. The raw count is therefore treated as **ambiguous** and
> excluded from scoring. Only the **resolution rate** is included as a positive
> indicator of service delivery quality.

---

## Domain 5 — Housing Vulnerability

Reflects the concentration of poverty and housing precarity within each
neighbourhood, using both the Market Basket Measure and rooming house
enforcement data.

| Indicator | Source | Direction |
|---|---|---|
| Poverty rate (population in MBM / total) | Higher Poverty Areas | ↓ Higher = worse |
| High poverty area designation | Higher Poverty Areas | ↓ Present = worse |
| Rooming house complaints per neighbourhood | Rooming House Enforcement | ↓ Higher = worse |
| Rooming house enforcement completion rate | Rooming House Enforcement | ↑ Higher = better |

---

## Domain 6 — Housing Market Pressure

Reflects the presence of short-term rental activity, particularly
investor-owned non-primary rentals which are associated with reduced housing
affordability.

| Indicator | Source | Direction |
|---|---|---|
| Short-term rentals per ward | Short-Term Rentals | Ambiguous — see note |
| % non-primary (investor-owned) STRs | Short-Term Rentals | ↓ Higher = worse |

> **Note:** Short-term rental data joins at the **electoral ward level** only,
> as no neighbourhood identifier is available in this dataset. Ward-level values
> are applied uniformly to all parcels within a ward, which introduces
> measurement imprecision at the neighbourhood level.

---

# 4. Composite Score Construction

The final WNLI score is computed in three steps:

**Step 1 — Normalize each indicator** to a 0–1 scale using min-max
normalization. Negative indicators (where higher raw values indicate worse
outcomes) are inverted so that **higher always means more liveable** across
all indicators.

**Step 2 — Average within domains.** Each domain score is the unweighted mean
of its constituent normalized indicators. Missing values are excluded from the
average rather than imputed, so neighbourhoods with fewer data sources still
receive partial scores.

**Step 3 — Average across domains.** The final composite score is the
unweighted mean of all six domain scores, multiplied by 100 to produce a
0–100 scale.

$$
\text{WNLI}_i = \frac{1}{6} \sum_{d=1}^{6} \text{Domain}_{d,i} \times 100
$$

Where $i$ indexes neighbourhoods and $d$ indexes the six domains.

---

# 5. Key Methodological Decisions

| Decision | Choice made | Rationale |
|---|---|---|
| **Index level** | Neighbourhood | Parcel-level data aggregated to neighbourhood before scoring |
| **Normalization** | Min-max (0–100) | Interpretable, bounded, easy to communicate |
| **Domain weights** | Equal (1/6 each) | Transparent starting point; sensitivity analysis will test alternatives |
| **Missing values** | Excluded from average | Avoids imputation bias; partial scores still produced |
| **Negative indicators** | Inverted before scoring | Higher always = more liveable across all indicators |
| **Time window** | 2021–2023 | Tightest common window across all event-level datasets |

---

# 6. Limitations

**Data availability.** Not all neighbourhoods have records in every dataset.
The Higher Poverty Areas dataset covers only ~60 designated high-poverty
neighbourhoods; Domain 5 scores will be missing for all other neighbourhoods
unless this is treated as "not high poverty" (a valid but explicit assumption).

**Neighbourhood name standardisation.** Several datasets use different naming
conventions for the same neighbourhood. Fuzzy matching and a manual remap table
were used to standardise keys before joining, but residual mismatches may remain.

**Ecological fallacy.** All scores are neighbourhood-level averages. They
describe the neighbourhood as a whole and should not be used to make inferences
about individual parcels or households.

**311 requests join failure.** At the time of writing, the 311 Requests dataset
has not successfully joined to the master due to neighbourhood key mismatches.
Domain 4 scores currently rely on trade permits only until this is resolved.

**Short-term rentals at ward level.** STR data is available at the electoral
ward level only. All neighbourhoods within a ward receive the same STR score,
which reduces sensitivity to within-ward variation.

---

# 7. Next Steps

1. Resolve outstanding neighbourhood name mismatches (particularly 311 Requests)
2. Validate domain scores against known neighbourhood characteristics
3. Conduct sensitivity analysis on domain weights
4. Map WNLI scores spatially using `centroid_lat` / `centroid_lon` from
   Assessment Parcels
5. Consult community stakeholders on whether equal domain weights reflect
   local priorities

---

# References

- Canadian Index of Wellbeing. (2016). *How are Canadians Really Doing?*
  University of Waterloo.
- Economist Intelligence Unit. (2023). *Global Liveability Index.*
- Statistics Canada. (2021). *Material and Social Deprivation Index.*
- United Nations Development Programme. (2023). *Human Development Index
  Technical Notes.*